---
title: |
  <center> Incomplete Data Analysis </center>
  <center> Assignment 2 </center>
author: "<center> Callum Abbott </center>"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(mice)
library(rjags)
library(JointAI)
```

All code used for this assignment can be found in the following repository [https://github.com/c-abbott/ida] under the folder `assignment-3`.

## Question 1
Consider the `nhanes` dataset in `mice`.

### Question 1a
*(a) What percentage of the cases are incomplete?*

**Solution:**
We observe the NHANES dataset has a total of $25$ cases, $12$ of which are
incomplete. This gives the incomplete percentage as $48.0\%$ to $3$ significant figures.

```{r echo=TRUE}
# Load in dataset
data(nhanes)
# Calculate NA percentage
na_percentage = sum(!complete.cases(nhanes)) / nrow(nhanes) * 100
```

### Question 1b
*(b) Impute the data with `mice` using the defaults with `seed=1`, in step 2 predict*
*`bmi` from `age`, `hyp`, and `chl` by the normal linear regression model, and then pool the*
*results. What are the proportions of variance due to the missing data for each parameter?*
*Which parameters appear to be most affected by the non-response?*

**Solution:**
Let *the proportions of variance due to the missing data for each parameter* be denoted
as $\lambda$. Mathematically, we specify $\lambda$ to be:
\begin{equation*}
  \lambda = \frac{B + \frac{B}{M}}{V^{\text{MI}}}
\end{equation*}

where $B$ denotes the *between-imputation* variance, $M$ denotes the number of 
imputed datasets by `mice` (default = 5) and $V^{\text{MI}}$ is the *total variance*.

Using the standard MICE procedure - `mice(), with(), pool()` - we yield the following
values for *the proportions of variance due to the missing data for each parameter* (3SF, seed=1):

* `age` = 0.686
* `hypten (yes)` = 0.350
* `chol` = 0.304

We observe that the variable `age` has the highest value of $\lambda$ indicating that
this variable is most affected by the non-response.
```{r echo=TRUE}
imps = mice(nhanes, seed=1, printFlag=FALSE) # MI step
fits = with(imps, glm(bmi ~ age + hyp + chl)) # Fitting lm to predict BMI
bmi_ests = pool(fits) # Gather useful summary statistics
summary(bmi_ests)
```

### Question 1c 
*(c) Repeat the analysis for* `seed` $∈ {2,3,4,5,6}$. *Do the conclusions remain the same?*

**Solution:**
Repeating an analogous analysis but now varying the seeds to the values given above we
yield different conclusions. For seeds 2, 3 and 6, the `age` variable had the largest
value of $\lambda$ whilst the `chl` and `hyp` variables had the largest $\lambda$ values
for seeds 4 and 5 respectively.

```{r}
bmi2 = pool(with(mice(nhanes, seed=2, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
bmi3 = pool(with(mice(nhanes, seed=3, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
bmi4 = pool(with(mice(nhanes, seed=4, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
bmi5 = pool(with(mice(nhanes, seed=5, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
bmi6 = pool(with(mice(nhanes, seed=6, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
```

### Question 1d
*(d) Repeat the analysis with M = 100 with the same seeds. Would you prefer*
*these analyses over those with M = 5? Explain why.*

**Solution:**
These results indicate that our conclusions are effectively worthless and are simply due
to random chance.

```{r}
mbmi1 = pool(with(mice(nhanes, seed=1, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
mbmi2 = pool(with(mice(nhanes, seed=2, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
mbmi3 = pool(with(mice(nhanes, seed=3, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
mbmi4 = pool(with(mice(nhanes, seed=4, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
mbmi5 = pool(with(mice(nhanes, seed=5, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
mbmi6 = pool(with(mice(nhanes, seed=6, m=100, printFlag=FALSE), glm(bmi ~ age + hyp + chl)))
```
